Higher-Order unification for free!

### 2024-04-21 paper/main.tex
re-add link-eta rule in constraint store
base -> basis (Yves)
update tex
minor nits

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $s$ such that $\exists \rho, \rho s$
  is an eta expansion.
\end{definition}

@@

The simple compiler described in this section serves as a base for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
using %the routine
\ref{clause:malloc}.

@@

When compiling a term $t$, we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems. Here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x,~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w_i \in W$, such that $w_i$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ is under the spine of binders
$x_1\ldots x_n$, $t$ can either be $x$ applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

@@

\noindent
Whenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
Invariant~\fullref{inv:linklhs} holds for this link. Moreover:

@@

\paragraph{Decompilation}
The decompilation of a \linketa is performed by unifying
the \lhs with the \rhs. Note that this unification never fails, since \lhs is a
flexible term not appearing in any other \linketa
(by \cref{def:progressetadedup}).

@@

\linketa{}s are meant to delay the unification of ``problematic'' terms until
we know for sure if the term has to be $\eta$-contracted or not.

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $t$, that is we run
    $\lambda x.X \appsep x\Ue{} t$
    \item if $X = \lambda x.t$ we run $X \Ue{} t$.
\end{enumerate}
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$: both terms are \wellb.
  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
  are both \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb, therefore, the substitution remains \wellb.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{This premise is fulfilled in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  succeeds. A $\hstep_i$ is made by a unification $u$ and a link
  progression $p$.
  Without loss of generality,\todo{explain why} we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$. $s_1$ is compiled to $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compiler to $t_2$ and $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft if $t_2$ is rigid.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}
          and $p$ unifies $\lambda x.t_1$ with $t_2$. Alternatively, if $s_1$ and $s_2$ fails
          to unify, it cannot be because of the head $\lambda$ constructors,
          since wither lam-lam or \ref{clause:eta1} move under it and find
          different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

@@

In \cref{sec:grounwork}, we introduced the definition of memory map \mapStore.
This map allows to decompile the \Ho substitution to the object language.
Given the higher-order nature of Coq the same unification
variable $X$ can be used multiple times with different arities.
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),a},
    {X,\lambda x.a}}}
  {{{A,a},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
In the unification problems \foUnifPb above, we see that $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected, breaking,
in turn \cref{prop:fidelity}. This incompatibility will only be detected during the
decompilation phase.
To address this issue, we post-processing the compiler's output we want to
ensure the follwoing property:

@@

\noindent
Note that the execution in \Ho{} may require allocating new variables as
explained in~\cref{sec:decompilation}. The property above does not
concern these.

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
\end{elpicode}
\todo{improve code}
%link-eta L R :- not (maybe-eta R), !, contract R S, L = S.

>>>>

\begin{definition}[\maybeeta]\label{def:maybeeta}
  \maybeeta is the set of terms $\lambda x.s$ such that $\exists \rho, \rho \lambda x.s$
  is an eta expansion. % without the lambda, 5.11 is not true, that
  % is we assume variables are in W
\end{definition}

@@

The simple compiler described in this section serves as a basis for the
extensions in \cref{sec:llam,sec:eta,sec:beta}.
Its main task is to beta normalize the term and map one syntax tree to
the other.
In order to bring back the substitution from \Ho{} to \Fo{} the compiler
builds a ``memory map'' connecting the \Fo{} variables to the \Ho ones
using %the routine
\ref{clause:malloc}.

@@

When compiling a term $t$, we need to determine if any
subterm $s \in \subterm{t}$ that is of the form $\lambda x. r$,
where $x$ occurs in $r$, can be a $\eta$-expansion, i.e. if
there exists a substitution $\rho$ such that $\rho (\lambda x.r) \Eo s$.
The detection of lambda abstractions that can ``disappear''
is not as trivial as it may seems. Here a few examples:
%
\begin{center}
  \begin{tabular}{lll}
    %Term & Status & Evidence \\\hline
    $\lambda x. f \appsep (A \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x ~\}$ \\
    $\lambda x. f \appsep (A \appsep x) \appsep x$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.a ~\}$\\
    $\lambda x. f \appsep x \appsep (A \appsep x)$ & $\not\in\maybeeta$ &\\
    $\lambda x. \lambda y. f \appsep (A \appsep x) \appsep (B \appsep y \appsep x)$ & $\in\maybeeta$ & $\rho = \{~ A \mapsto \lambda x.x~;~ B \mapsto \lambda y.\lambda x.y ~\}$
  \end{tabular}
\end{center}
\vspace{4pt}

@@

\begin{lemma}\label{lem:reduceto}
A $\beta$-normal term $s = \lambda x_1 \ldots x_n.t$
%, where $x, x_1\ldots x_n$ can occur in $t$, 
\reduceto{} $x$ only if one of the following three conditions holds:
\begin{enumerate}
  \item $n = 0$ and $t = x$;
  \item $t$ is the application of $x$ to a list
     of terms $l$ and each $l_i$ \reduceto{} $x_i$
     (e.g. $\lambda x_1 \ldots x_n.x \appsep x_1 \ldots x_n \Eo{} x$) ;
  \item $t$ is a unification variable with scope
    $W$, and for any $v \in \{ x, x_1 \ldots x_n \}$,
    there exists a $w \in W$, such that $w$ \reduceto{} $v$
    (if $n = 0$ this is equivalent to $x \in W$).
\end{enumerate}
\end{lemma}
\begin{proof}[Proof sketch]
Since our terms are in $\beta$-normal form there is only
one rule that can play a role (namely \ref{clause:eta1}),
hence if the term $s$ is not exactly $x$ (case 1)
it can only be an $\eta$-expansion of $x$, or a unification
variable that can be assigned to $x$, or a combination of both.
If $s$ begins with a lambda, then the lambda can only disappear by $\eta$
contraction. In that case, the term $t$ is under the spine of binders
$x_1\ldots x_n$ can either be $x$ applied to terms that can
\reduceto{} these variables (case 2), or a unification variable
that can be assigned to that application (case 3).
\end{proof}

@@

\noindent
Whenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1} and is connected to \elpiIn{lam F1} via a \linketa.
Invariant~\fullref{inv:linklhs} holds for this link. Moreover:

@@

\paragraph{Decompilation}
The decompilation of a \linketa is performed by unifying
the \lhs with the \rhs. Note that this unification never fails, since the \lhs is a
flexible term not appearing in any other \linketa
(by \cref{def:progressetadedup}).

@@

\linketa{}s are meant to delay the unification of ``problematic'' terms until
we know for sure if their head lambdas can be $\eta$-contracted or not.

\newcommand{\progressetaleft}{\emph{$\eta$-progress-\lhs}\xspace}
\begin{definition}[\progressetaleft]\label{def:progressetaleft}
A link \linketaM{\Gamma}{X}{t} is removed from \linkStore when
$X$ becomes rigid. Let $y\in\Gamma$, there are two cases:
\begin{enumerate}
  \item if $X = a$ or $X = y$ or $X = f \appsep a_1\ldots a_n$
    we unify the $\eta$-expansion of $X$ with $t$, that is we run
    $\lambda x.X \appsep x\Ue{} t$
    \item if $X = \lambda x.s$ we run $X \Ue{} t$.
\end{enumerate}
\end{definition}

@@

\begin{proof}[Proof sketch]
  Let $\sigma$ be the substitution, such that $\wellb(\sigma X)$ (since \Ue preserves \wellb).
  By \cref{inv:link-eta-right}, we have $\wellb(t)$.
  If $\sigma X = \lambda x.r$, then $r$ is unified with $t$ and both terms are in \wellb.
  Otherwise, $\lambda x. X x$ is unified with $\lambda x. t$, and $X x$ and $t$
  are both in \wellb.
\end{proof}

@@

\begin{proof}[Proof sketch]
  By \cref{lemma:unif-eta-1,lemma:unif-eta-3}, every
  unification performed by the activation of a \linketa is done between
  terms in \wellb.
\end{proof}

@@

\begin{theorem}[Fidelity in $\wellb_{\beta \eta}$]
  % F x = z, F x y = z y
  Given a list of unification problems \foUnifPb, such that
  $\wellb_{\beta \eta}(\foUnifPb)$, if the 
  memory map is bijective, then
  the introduction 
  of \linketa guarantees \fullref{prop:fidelity}.
  \footnote{The bijective property of \mapStore is proved in \cref{sec:invariant1}}
  \label{lemma:fidelity-maybeeta}
\end{theorem}

\begin{proof}[Proof sketch]\todo{to reformulate}
  % For any step $i$, we want to prove that $\fstep_i$ succeeds iff $\hstep_i$
  % succeeds. A $\hstep_i$ is made by a unification
  % $u = \sigma_{i-1}\hoUnifPb_{i_l} \Ue \sigma_{i-1}\hoUnifPb_{i_r} \mapsto \sigma_i'$ and a link
  % progression $p = \progress(\linkStore_{i-1},\sigma_i') \mapsto (\linkStore_i,\sigma_i)$.
  We want to prove that \fstep succeeds iff \hstep
  succeeds, where \hstep is made by a unification step $u$ and a link
  progression $p$.
  Without loss of generality we consider a
  \Fo unification problem of the form $s_1 \Uo s_2$
  where $s_1 \in \maybeeta$ and is compiled to a variable $X$ with the accessory link
  $\linketaM{\Gamma}{X}{\lambda x.t_1}$. 
  The unification $u$ always succeeds, since $X$ is a fresh variable, we 
  only have to consider the link progression $p$.
  Two cases should be analyzed:
  \begin{itemize}
    \item $s_2 \notin \maybeeta$,
          in this case $s_2$ is compilet to $t_2$ and the unification step $u$ assigns $t_2$ to $X$
          and triggers \progressetaleft if $t_2$ is rigid.
          If $s_1$ and $s_2$ unify thanks to \ref{clause:eta1} then
          progression $p$ unifies $\lambda x.t_1$ with the $\eta$-expansion of $t_2$ namly
          $\lambda x.t_2\appsep x$, otherwise they unify thanks to \cref{clause:lam-lam}\todo{name}
          and the link progress $p$ unifies $\lambda x.t_1$ with $t_2$.\\ Alternatively, if $s_1$ and $s_2$ fail
          to unify, it cannot be because of the head $\lambda$ constructors,
          since with \cref{lam-lam}\todo{name} or \ref{clause:eta1} move under it and find
          different sub terms $s_1'$ and $s_2'$. By \cref{th:fidelity-beta},
          \Ue fails on terms $t_1'$ and $t_2'$ even if \progressetaleft
          $\eta$-expands $t_2$.

@@

We want to allow for partial application of functions.
As a consequence
the same unification variable $X$ can be used multiple times
with different arities.
Then it is the case the
memory map \mapStore generated by the compiler is not
a bijection. For example:
\begin{comment}
%
% make test ONLY=7001 TEX=tex
\printAlll
  {{{\lambda x.\lambda y.(X\appsep y\appsep x),\lambda x.\lambda y.x},
    {\lambda x.(f\appsep (X\appsep x)\appsep x),Y}}}
  {{{A,\lambda x.\lambda y.x},
    {D,F}}}
  {{{X,E,1},
    {Y,F,0},
    {X,C,2}}}
  {{{\eta,,D,\lambda x.(f\appsep E_{x}\appsep x)},
    {\eta,,A,\lambda x.B_{x}},
    {\eta,x,B_{x},\lambda y.C_{y x}}}}
\end{comment}
% make test ONLY=7007 TEX=tex
\printAlll
  {{{\lambda x.(X\appsep x),a},
    {X,\lambda x.a}}}
  {{{A,a},
    {C,\lambda x.a}}}
  {{{X,C,0},
    {X,B,1}}}
  {{{\eta,,A,\lambda x.B_{x}}}}
% \todo{simplify example since no prog right}
in the unification problems \foUnifPb above, $X$ is used with arity $1$ in
$\foUnifPb_1$ and with arity $0$ in $\foUnifPb_2$. 
In order to preserve \fullref{inv:uvaarity} the compiler
did generate two entries in \mapStore for $X$, namely $B$ and $C$. 
However, there is no connection between these two variables
and any incompatible assignements to them would not be detected, breaking,
in turn \cref{prop:fidelity}. 
% This incompatibility will only be detected during the
decompilation phase.
To address this issue, we post-process the compiler's output
to ensure the follwoing property:

@@

\noindent
Note that the execution in \Ho{} may require allocating new variables as
explained in~\cref{sec:decompilation}. The property above does not
concern these variables since they are fresh and have no corresponding
variable in \Fo.

@@

\begin{elpicode}
link-eta L R :- var L, maybe-eta R HolesInR, !,
  declare_constraint (link-eta L R) [L|HolesInR].
link-eta L R :- not (var L), !, eta-progress-lhs L R.
\end{elpicode}
% \todo{improve code}
% link-eta L R :- not (maybe-eta R), !, contract R S, L = S.
